<!-- Calculate quartiles for text -->

```{r}
library(here)
source(here("R/calc_quartiles.R"))
```

# Economic cost-effectiveness {#sec-economic}

**Gate-to-gate ATFM delays in 2024 reached 29.6 million minutes, resulting in an estimated cost of €3.9 billion on airspace users, the highest level ever recorded. This figure represents more than one-third of ATM/CNS provision costs at the Pan-European system level and surpasses total ATCO in OPS employment costs.**

The concept of economic cost-effectiveness, developed by the PRC, is defined as the sum of gate-to-gate ATM/CNS provision costs and the costs of ground ATFM delays for both en‐route and airport, all expressed per composite flight-hour. This economic performance indicator is meant to capture trade‐offs between quality of service provided and costs[^2].

[^2]: See <https://ansperformance.eu/economics/ace/ace-handbook/> for more information on the methodology used to compute composite flight-hours and economic costs.

@fig-figure-3-1 shows [preliminary results]{.underline} on the changes in economic cost-effectiveness over {{< var doc.year_report_minus5 >}} - {{< var doc.year_report >}} at Pan-European system level. The left-hand side shows the changes in unit economic costs, while the chart on the right-hand side provides complementary information on the year-on-year changes in ATM/CNS provision costs, composite flight-hours and unit costs of ATFM delays. Unit economic costs rose by +2.4% in 2024, driven by a +14.4% increase in the unit costs of ATFM delays, which outweighed a -1.4% reduction in ATM/CNS provision costs per composite flight-hour. As a result, unit economic costs were +4.7% above 2019 levels.

In 2024, ATFM delays increased by +20.7% to some 29.6M minutes, raising the share of ATFM delays in unit economic costs to 26.9% (from 24.1% in 2023). This is the highest share ever recorded. Considering the total estimated costs of ATFM delays in 2024 (€3.9 billion), it represents more than one-third of ATM/CNS provision costs at the Pan-European system level and surpasses total ATCO in OPS employment costs.

:::: {#fig-figure-3-1 layout-ncol="2"}
::::: {#fig-figure-3-1-left}
```{r}
#| file: R/hlsr_evo_eco_ce.R
#| out.height: !expr if (knitr::is_html_output()) "300px" else "245px"
#| out.width: "100%"
```

 
:::::

::::: {#fig-figure-3-1-right}
```{r}
#| file: R/hlsr_evo_eco_ce_comp.R
#| out.height: !expr if (knitr::is_html_output()) "300px" else "245px"
#| out.width: "100%"
```

 
:::::

Trend of unit economic costs at Pan-European system level, {{< var doc.year_report_minus5 >}}-{{< var doc.year_report >}} (real terms)
::::

@fig-figure-3-2 shows [preliminary results]{.underline} at ANSP level (dotted lines represent the 1st and 3rd quartiles, €`r q1_3_2` and €`r q3_3_2`, respectively).

::: {#fig-figure-3-2}
```{r}
#| file: R/hlsr_eco_ce.R
#| format:
#|   html:
#|     out.height: "490px"
#|     out.width: "100%"
```

Economic gate-to-gate cost-effectiveness, {{< var doc.year_report >}} 
:::

For 15 ANSPs ATFM delays represented more than 20% of their unit economic costs (see ANSPs with the largest red and yellow portions, e.g. HungaroControl, Croatia Control, HASP, M-NAV, BHANSA, etc.). In absolute terms (cumulative ATFM delays in minutes) DSNA, DFS, ENAIRE, HungaroControl and NATS were the five ANSPs generating the highest levels of ATFM delays in 2024 (61.8% of the Pan-European system total ATFM delays).

The primary sources of ATFM delays in 2024 were capacity and staffing shortages, followed by adverse weather conditions. The situation at HungaroControl was particularly exceptional, as a combination of factors led to unusually high levels of ATFM delays. According to the Network Manager’s Operations Report[^3], several issues impacted HungaroControl performance in 2024, most notably unforeseen staffing challenges and a significant surge in traffic demand (an increase of +6% compared to 2023 and +15% compared to 2019), partly driven by the situation in Ukraine. Adverse weather further compounded these difficulties. It is noteworthy that the situation has improved significantly in 2025.

Further analysis of the relationship between changes in ANSPs costs, traffic and unit costs will be analysed in detail in the forthcoming ACE report.

[^3]: Annual Network Operations Report 2024, available at <https://www.eurocontrol.int/publication/annual-network-operations-report-2024>].
