# Economic cost-effectiveness {#sec-economic}

The concept of economic cost-effectiveness, developed by the PRC, is defined as the sum of gate-to-gate ATM/CNS provision costs and the costs of ground ATFM delays for both en‐route and airport, all expressed per composite flight-hour. This economic performance indicator is meant to capture trade‐offs between ATC capacity and costs[^economic-cost-effectiveness-1].

[^economic-cost-effectiveness-1]: See https://ansperformance.eu/economics/ace/ace-handbook/ for more information on the methodology used to compute composite flight-hours and economic costs.

@fig-figure-3-1 shows [preliminary results]{.underline} on the changes in economic cost-effectiveness over {{< var doc.year_report_minus5 >}} - {{< var doc.year_report >}} at Pan-European system level. @fig-figure-3-1-left shows the changes in unit economic costs, while @fig-figure-3-1-right provides complementary information on the year-on-year changes in ATM/CNS provision costs, composite flight-hours and unit costs of ATFM delays. Unit economic costs significantly reduced in {{< var doc.year_report >}} (-24.8%) but their level remains substantially higher than before the COVID-19 pandemic.

::: {#fig-figure-3-1 layout-ncol="2"}
::: {#fig-figure-3-1-left}

```{r}
#| file: R/hlsr_evo_eco_ce.R
#| out.height: "300px"
#| out.width: "100%"
```
&nbsp;
:::

::: {#fig-figure-3-1-right}
```{r}
#| file: R/hlsr_evo_eco_ce_comp.R
#| out.height: "300px"
#| out.width: "100%"
```
&nbsp;
:::
Trend of unit economic costs at Pan-European system level, {{< var doc.year_report_minus5 >}}-{{< var doc.year_report >}} (real terms)
:::

@fig-figure-3-2 shows [preliminary results]{.underline} at ANSP level (dotted lines represent the 1st and 3rd quartiles). On average, the share of ATFM delays in {{< var doc.year_report >}} was some 4% (compared to 3% in {{< var doc.year_report_minus1 >}} and 22% in {{< var doc.year_report_minus2 >}}), and only four ANSPs had ATFM delays representing 5% or more of their unit economic costs: HASP (32%), NAV Portugal (8%), DSNA (7%) and LVNL (5%).

::: {#fig-figure-3-2}
```{r}
#| file: R/hlsr_eco_ce.R
#| out.height: "450px"
#| out.width: "100%"
```
Economic gate-to-gate cost-effectiveness, {{< var doc.year_report >}}
:::


